---
title: ggplot2 Avançado
author: Claudia Tanaka (claudia.tanaka@ans.gov.br)
date: "Atualizado em `r format(Sys.time(), '%d/%m/%Y')`"

# params: 
#   path.in: _pasta.de.entrada/
#   path.out: _pasta.de.saida/

output:
  html_notebook:
    toc: yes
    toc_depth: 4
    toc_float: yes
    code_folding: hide
---

*Desenvolvido em R 4.2.1 Patched e RStudio 2022.07.0+548 for Windows*

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = TRUE )

library(tidyverse)
library(lubridate)
library(patchwork)
```

<br>

Theme settings

```{r}
theme_set(theme_light())
theme_update(
  # line = element_line(colour = 'gray70'),
  # rect = element_rect(colour = 'gray70'),
  # text = element_text(colour = 'gray70'),
  
  panel.grid.minor = element_blank(),
  # panel.border = element_rect(fill = NA, color = "gray90"),
  # 
  # axis.title = element_text(),
  # axis.ticks = element_blank(),
  # axis.text = element_text(colour = "gray50", size = 9),
  # 
  # plot.title = element_text(size = 12, colour = "gray30", face = "bold"),
  # plot.subtitle = element_text(face = 'italic', colour = "gray50", size = 10),
  plot.caption = element_text(colour = "gray50", hjust=0, size = 8),

  legend.title = element_blank(),
  # legend.position = "top",
  # legend.text = element_text(color="Gray50", size = 8),
  # plot.margin = margin(rep(10, 4)) # add some white space around the plot
)
```


<br>

# Correlation plot

```{r fig.width=8, fig.height=8}
library(GGally)

iris |>
  ggpairs(
    columns=1:4,
    aes(colour=Species)
  ) +
  theme(panel.grid = element_line(linetype = "dotted"))
```



<br>

# Radar plot

The data should be organized as follows:

- Row 1 must contain the maximum values for each variable

- Row 2 must contain the minimum values for each variable

- Data for cases or individuals should be given starting from row 3

- The number of columns or variables must be more than 2.

```{r fig.asp=1, fig.width=10}
library(fmsb)

df.spider <- rbind(
  mpg |> 
    summarise(across(c(where(is.numeric), -year), min)) |>
    mutate(name = "min"),
  mpg |> 
    summarise(across(c(where(is.numeric), -year), max)) |>
    mutate(name = "max"),
  # Calcula as medianas de cada classe de veículo
  mpg |> 
    summarise(across(c(where(is.numeric), -year), median), .by=class) |>
    rename(name=class)
) |> column_to_rownames("name")


# Esse plot automaticamente faz rescale das variáveis para min=0 e max=1
radarchart(
  df.spider[1:5,],
  # The grid
  axistype=1,          # labels the grid intervals
  seg=2,               # qty of grid intervals
  axislabcol="gray40", # color of the grid labels
  cglcol="grey",       # grid line color see ? for line width, linetype
  # Variable labels
  vlabels=c("Displacement", "Cylinders", "City\nmpg", "Highway\nmpg"), # label the variables
  vlcex=1,  # var labels size
  # POLYGON
  pcol=RColorBrewer::brewer.pal(7, "Dark2"), 
  plwd=2,   # polygon line width
  plty=1,   # polygon line types
  na.itp=F, # como interpolar nulos
  title = "MPG Median comparison by Vehicle Class"
)
```


```{r fig.asp=1}
# Reduce plot margin using par()
op <- par(mar = c(2,2,2,2))

radarchart(
  df.spider[1:5,],
  # The grid
  axistype=1,          # type of grid labels
  seg=5,               # qty of internal grid interval labels
  caxislabels=c("baixo", "", "", "", "alto"), # customize internal grid interval labels
  axislabcol="gray40", # color of the grid labels
  cglcol="grey",       # grid line color see ? for line width, linetype
  # Variable labels
  vlabels=c("Displacement", "Cylinders", "City mpg", "Highway\nmpg"), # label the variables
  vlcex=.7,  # var labels size
  # POLYGON
  pcol=viridis::viridis(3),
  pfcol=scales::alpha(viridis::viridis(3), 0.1),
  plwd=2,   # polygon line width
  plty=1,   # polygon line types
  na.itp=F, # como interpolar nulos
  title = "MPG Median comparison by Vehicle Class"
)

# Add an horizontal legend
legend(
  x="bottom", legend=rownames(df.spider[-c(1,2),]), horiz=TRUE,
  bty="n", pch=20 , col=viridis::viridis(7),
  text.col="black", cex=1, pt.cex=1.5
)

par(op)
```


```{r fig.asp=.2, fig.width=9}
colors <- RColorBrewer::brewer.pal(7, "Dark2")
titles <- rownames(df.spider[-c(1,2),])

# Reduce plot margin using par() & split screen in 7 parts
op <- par(mar = c(1,1,1,1))
par(mfrow=c(1,7))

for (i in 1:7) {
  radarchart(
    df.spider[c(1,2,i+2),],
    # The grid
    axistype=1,          # type of grid labels
    seg=5,               # qty of internal grid interval labels
    caxislabels=c("baixo", "", "", "", "alto"), # customize internal grid interval labels
    axislabcol="gray40", # color of the grid labels
    cglcol="grey",       # grid line color see ? for line width, linetype
    # Variable labels
    vlabels=c("Displacement", "Cylinders", "City mpg", "Highway\nmpg"), # label the variables
    vlcex=.7,  # var labels size
    # POLYGON
    pcol=colors[i],
    pfcol=scales::alpha(colors[i], 0.05),
    plwd=2,   # polygon line width
    plty=1,   # polygon line types
    na.itp=F, # como interpolar nulos
    title = titles[i]
  )
}
par(op)
```


<br>

# Lollipop plot

<br>

# Waterfall chart


<br>

# DISTRIBUTIONS


<br>

### Histogram

[Advanced histogram features](https://ggplot2.tidyverse.org/reference/geom_histogram.html)


```{r}
# Data
set.seed(1234)
df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5),
                 rnorm(200, mean=65, sd=5)))
  )

# Calculate the mean of each group:
mu <- df |> group_by(sex) |> summarise(grp.mean = mean(weight))
```



```{r fig.width = 5, fig.height = 3}
ggplot(df, aes(x=weight)) + 
  geom_histogram(binwidth=1) + # Control binwidth= OR bins=
  geom_histogram(color="black", fill="white") + # Change colors
  geom_vline(aes(xintercept=mean(weight)), color= "blue", linetype="dashed", size=1)
```

<br>

Histogram by group

```{r fig.width = 5, fig.height = 3}
ggplot(df, aes(x=weight, fill=sex, color=sex)) + # Change line and fill colors by group
  geom_histogram(alpha=0.5, position="dodge") + # Interleaved histograms
  # add mean lines
  geom_vline(data=mu, aes(xintercept=grp.mean, color=sex),
             linetype="dashed")
```

<br>

### Density plot

<br>

```{r fig.width = 5, fig.height = 3}
 ggplot(df, aes(x=weight)) + 
  geom_density(color="darkblue", fill="lightblue", linetype="dashed") + # Format line and fill
  # add mean line
  geom_vline(aes(xintercept=mean(weight)),
             color="blue", linetype="dashed", size=1)
```

<br>

Density plot by group

<br>

Change line colors. Density plot line colors can be automatically controlled by the levels of sex :

```{r fig.width = 5, fig.height = 3}
ggplot(df, aes(x=weight, color=sex, fill=sex)) +
  geom_density(alpha=0.4)+
  # Add mean lines
  geom_vline(data=mu, aes(xintercept=grp.mean, color=sex),
             linetype="dashed")
```



```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```


<br>

### Histogram + Density

The histogram is plotted with density instead of count values on y-axis in order to be on the same scale as the density plot. Overlay with transparent density plot.

```{r fig.width = 5, fig.height = 3}
# Data
set.seed(1234)
df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5),
                 rnorm(200, mean=65, sd=5)))
  )

# Calculate the mean of each group:
mu <- df |> group_by(sex) |> summarise(grp.mean=mean(weight))


ggplot(df, aes(x=weight)) + 
 geom_histogram(aes(y=after_stat(density)), colour="black", fill="white", bins=30)+
 geom_density(alpha=.2, fill="#FF6666") 
```


<br>

Combine by group

```{r fig.width = 5, fig.height = 3}
# Color by groups
ggplot(df, aes(x=weight, color=sex, fill=sex)) + 
 geom_histogram(aes(y=..density..), alpha=0.5, 
                position="identity")+
 geom_density(alpha=.2) 
```

<br>

### Cumulative frequency


```{r fig.width = 6, fig.height = 5}
ggplot(df, aes(x=weight, color=sex, fill=sex)) + 
  stat_ecdf() +
  labs(
    title = "Distribuição da Frequência Acumulada",
    y = "Empirical Cumulative Density Function (ecdf)"
  )
```




```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```




<br>

### Box plot


```{r}
set.seed(2021)

data <- tibble(
  group1 = factor(c(rep("Group A", 100), rep("Group B", 250), rep("Group C", 25))),
  group2 = factor(c(rep(c("Group X","Group Y"), 187), "Group Y")),
  value = c(
    seq(0, 20, length.out = 100),
    c(rep(0, 5), 
      rnorm(30, 2, .1), rnorm(90, 5.4, .1), rnorm(90, 14.6, .1), rnorm(30, 18, .1), 
      rep(20, 5)),
    rep(seq(0, 20, length.out = 5), 5))
  ) |> 
  rowwise() |> mutate(value = if_else(group1=="Group B", value + rnorm(1, 0, .4), value))
```


```{r fig.width = 5, fig.height = 3}
# grouped boxplot
ggplot(data, aes(x=group1, y=value, fill=group2)) + geom_boxplot() + 
  theme(legend.title = element_text())
```


<br>

Box plots are great! Box plots are an artwork combining many summary statistics into one chart type. But in my opinion they are not always helpful1. They also have a high potential of misleading your audience—and yourself. Why?

So tell me: How big is the sample size? Are there underlying patterns in the data? Difficult?

Sure, adding a note on the sample size might be considered good practice but it still doesn’t tell you much about the actual pattern.



```{r fig.width = 5, fig.height = 3}
## function to return median and labels
n_fun <- function(x){
  return(data.frame(y = median(x) * .9, 
                    label = paste0("n = ",length(x))))
}

ggplot(data, aes(x = group1, y = value)) +
  geom_boxplot(fill = "grey92") +
  ## use summary function to add text labels
  stat_summary(
    geom = "text",
    fun.data = n_fun,
    size = 3
  )
```

<br>

Esse truque também funciona com facets

```{r fig.asp=.5}
## function to return median and labels
n_fun <- function(x){
  return(data.frame(y = median(x) * .9, 
                    label = paste0("n = ",length(x))))
}

ggplot(data, aes(x = group2, y = value)) +
  geom_boxplot(fill = "grey92") +
  facet_wrap(~group1) +
  ## use summary function to add text labels
  stat_summary(
    geom = "text",
    fun.data = n_fun,
    size = 3
  )
```


<br>

**We Can Do Better: Add Raw Data**   

An obvious improvement is to add the data points. Since we know already that Group 2 consists of 200 observations, let’s use jitter strips instead of a traditional strip plot:

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) +
  geom_boxplot(fill = "grey92") +
  ## use either geom_point() or geom_jitter()
  geom_point(
    ## draw bigger points
    size = 2,
    ## add some transparency
    alpha = .3,
    ## add some jittering
    position = position_jitter(
      ## control randomness and range of jitter
      seed = 1, width = .2
    )
  )
```

<br>

Oh, the patterns are very different! Values of Group 1 are uniformly distributed. The values in Group 2 are clustered with a distinct gap around the group’s median! And the few observations of Group 3 are all integers.

We can improve the look a bit further by plotting the raw data points according to their distribution 

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) +
  geom_boxplot(fill = "grey92") +
  ggforce::geom_sina(
    ## draw bigger points
    size = 1.5,
    ## add some transparency
    alpha = .4,
    ## control range of the sina plot
    maxwidth = .8
  ) 
```

<br>

Alternatively, we may be interested in specifically highlighting the outliers...

```{r fig.width = 6, fig.height = 4}
df.outliers <- mpg |> 
  mutate(
    cyl = as.factor(cyl),
    outlier.color = case_when(
      hwy > (quantile(hwy, .75) + 1.5 * IQR(hwy)) ~ "red",
      hwy < (quantile(hwy, .25) - 1.5 * IQR(hwy)) ~ "blue",
      .default = NA
    ), 
    .by = cyl
  ) 

df.outliers |> 
  ggplot(aes(x = cyl, y = hwy)) +
  geom_boxplot(fill = "grey92", outlier.color = NA) +
  geom_jitter(color = df.outliers$outlier.color, width = .3, alpha=.4, size=2.5)
```

<br>

Boxplot com STATS (limites) customizados. Especialmente útil para distribuições "skewed" (não normais) como Preço, Reajuste:

```{r}
# SLOWER METHOD:
p1 <- ggplot(iris, aes(x = Species, xend=Species, y = Petal.Width)) +
  # Add fences (outlier limits)
  stat_summary(geom="segment", fun.data = function(x) {
    data.frame(
      y = quantile(x, 0.01),
      yend = quantile(x, 0.99)
    )}, arrow=arrow(angle=90, ends="both", length=unit(0.2, "cm"))) +
  # Add boxplot with customized stats
  stat_summary(geom="boxplot", fun.data=function(x) {
    data.frame(
      ymin   = quantile(x, 0.01),
      lower  = quantile(x, 0.25),
      middle = median(x, 0.5),
      upper  = quantile(x, 0.75),
      ymax   = quantile(x, 0.99) 
    )}, outlier.shape = NA, width=0.4) +
  labs(title="Slower method with less code lines")



# FASTER METHOD:

# Function to calculate boxplot stats
bp.stats = function (x, probs = c(0.01, 0.25, 0.5, 0.75, 0.99)) {
  r <- quantile(x, probs = probs, na.rm = TRUE)
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}

# Function to calculate fence limits for outliers
segment.stat = function(x, probs = c(0.01, 0.99)) {
  r <- unname(quantile(x,probs))
  r <- c(y=r[1], yend=r[2])
  r
}

p2 <- ggplot(iris, aes(x=Species, xend=Species,  y=Petal.Width)) +
  # Add fences (outlier limits)
  stat_summary(fun.data=segment.stat, geom="segment",
               arrow=arrow(angle=90, ends="both", length=unit(0.2, "cm"))) +
  # Add boxplots
  stat_summary(fun.data=bp.stats, geom="boxplot", width=0.4) +
  labs(title="Faster method with more code lines")

p1+p2
```

<br>

Boxplots for Skewed Distributions using robust measurement for limits (medcouple)

Hubert M., Vandervieren E., 2008. "An adjusted boxplot for skewed distributions", Computational
Statistics and Data Analysis 52, 5186-5201.

```{r}
data <- rbeta(10000, 10, 1) |> as_tibble() |> 
  mutate(ol.tukey = if_else(value<(quantile(value,.25)-1.5*IQR(value)) | 
                              value>quantile(value,.75)+1.5*IQR(value), "red", NA))

hist(data$value,xlab="rbeta()",main="Histogram of rbeta()")
```


<br>

```{r}
# Function to calculate High Density Interval
hdi.stat = function(x, credMass=0.5) {
  r = unname(HDInterval::hdi(x, credMass=credMass))
  r = c(y=r[1], yend=r[2])
  r
}

# SLOWER METHOD:
p1 <- ggplot(data, aes(y=value, x="rbeta()", xend="rbeta()")) +
  # Add fences (outlier limits)
  stat_summary(geom="segment", fun.data = function(x) {
    data.frame(
      y = quantile(x,.25)-1.5*IQR(x),
      yend = quantile(x,.75)+1.5*IQR(x)
    )}, arrow=arrow(angle=90, ends="both", length=unit(0.2, "cm"))) +
  # Add boxplot 
  geom_boxplot(width=.4, outlier.color=NA) +
  geom_jitter(color = data$ol.tukey, width=.2, alpha=.2, size=2.5, shape=1) +
  # Add HDI
  stat_summary(fun.data=hdi.stat, geom="segment", 
               position=position_nudge(x=-0.3), colour="orange", size=.9,
               arrow=arrow(angle=90, ends="both", length=unit(0.1, "cm"))) +
  labs(title="Tukey BoxPlot with High Density Interval in orange") +
  coord_flip()


# Function to calculate boxplot stats
bp.stats = function (x) {
  r <- robustbase::adjboxStats(x)$stats
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}
# Function to calculate fence limits for outliers
segment.stat = function(x) {
  r <- robustbase::adjboxStats(x)$stats[c(1,5)]
  r <- c(y=r[1], yend=r[2])
  r
}

ol.adjbp <- robustbase::adjboxStats(data$value)$out

data <- data |> 
  mutate(ol.adjbp = if_else(value %in% ol.adjbp, "red", NA))

p3 <- ggplot(data, aes(y=value, x="rbeta()")) +
  # Add fences (outlier limits)
  stat_summary(aes(xend="rbeta()"), geom="segment", fun.data = segment.stat, 
               arrow=arrow(angle=90, ends="both", length=unit(0.2, "cm"))) +
  # Add boxplot
  stat_summary(fun.data=bp.stats, geom="boxplot", width=0.4, outlier.color="red") +
  geom_jitter(color = data$ol.adjbp, width=.2, alpha=.4, size=2.5, shape=1) +
  labs(title="Adjusted BoxPlot for Skewed Distributions") +
  coord_flip()

suppressWarnings(print(p1/p3))
```



<br>

### Violin plots

Violin plots can be used to visualize the distribution of numeric variables. It’s basically a mirrored density curve, representing the number of data points along a continuous axis.

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) + geom_violin(fill = "grey92")
```

<br>

By default, the violin plot can look a bit odd. The default setting (scale = "area") is misleading. Group 1 looks almost the same as Group 3, while consisting of four times as many observations. Also, the default standard deviation of the smoothing kernel is not optimal in our case since it hides the true pattern by smoothing out areas without any data.

We can manipulate both defaults by setting the width to the number of observations and adjusting the bandwidth (bw). Aesthetically, I prefer to remove the default outline:

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) +
  geom_violin(
    fill = "grey72", 
    ## remove outline
    color = NA, 
    ## width of violins mapped to # observations
    scale = "count", 
    ## custom bandwidth (smoothing)
    bw = .4
  )
```

<br>

The violin plot allows an explicit representation of the distribution but doesn’t provide summary statistics. To get the best of both worlds, it is often mixed with a box plot—either a complete box plot with whiskers and outliers or only the box indicating the median and interquartile range (IQR):

```{r fig.width = 8, fig.height = 5}
p1 <- ggplot(data, aes(x = group1, y = value)) +
  geom_violin(
    fill = "grey72", 
    color = NA, 
    scale = "count", 
    bw = .5
  ) +
  geom_boxplot(
    ## remove white filling
    fill = NA, 
    ## reduce width
    width = .1
  )

p2 <- ggplot(data, aes(x = group1, y = value)) +
  geom_violin(
    fill = "grey72", 
    color = NA, 
    scale = "count", 
    bw = .5
  ) +
  geom_boxplot(
    ## remove white filling
    fill = NA, 
    ## reduce width
    width = .1,
    ## remove whiskers
    coef = 0, 
    ## remove outliers
    outlier.color = NA ## `outlier.shape = NA` works as well
  )

p1+p2
```

<br>

You might wonder: why should you use violins instead of box plots with superimposed raw observations? Well, in case you have many more observations, all approaches of plotting raw data become difficult. In terms of readability as well as in terms of computation(al time). Violin plots are a good alternative in such a case, and even better in combination with some summary stats. But we also can combine all three…




### Raincloud

<br>

[Visualizing distributions with Raincloud Plots in ggplot2](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-and-how-to-create-them-with-ggplot2/)

Raincloud plots can be used to visualize raw data, the distribution of the data, and key summary statistics at the same time. Actually, it is a hybrid plot consisting of a halved violin plot, a box plot, and the raw data as some kind of scatter.

Here, I combine two layers from the `{ggdist}` package, namely `stat_dots()` to draw the rain and `stat_halfeye()` to draw the cloud. Both are plotted with some justification to place them next to each other and make room for the box plot. I also remove the slab interval from the halfeye by setting `.width = 0` and `point_colour = NA`. The plot needs some manual styling and the values for justification and the number of bins depends a lot on the data. To get rid of the white space on the left, we simply add a limit the x axis.

```{r fig.width = 5, fig.height=3}
ggplot(data, aes(x = group1, y = value)) + 
  ## add half-violin from {ggdist} package
  ggdist::stat_halfeye(
    ## custom bandwidth
    adjust = .5, 
    ## adjust height
    width = .6, 
    ## move geom to the right
    justification = -.2, 
    ## remove slab interval
    .width = 0, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .12, 
    ## remove outliers
    outlier.color = NA ## `outlier.shape = NA` works as well
  ) +
  ## add dot plots from {ggdist} package
  ggdist::stat_dots(
    ## orientation to the left
    side = "left", 
    ## move geom to the left
    justification = 1.1, 
    ## adjust grouping (binning) of observations 
    binwidth = .25
  ) + 
  ## remove white space on the left
  coord_cartesian(xlim = c(1.2, NA)) #+ coord_flip()
```

<br>


One can also solely rely on layers from the `{ggdist}` package by using the default halfeye which consists of a density curve and a slab interval.

But...

While I love the reduced design and the possibility to indicate two different ranges (here the **interquartile range** and the **95% quantile**) I admit that this alternative is less intuitive and potentially even misleading since they look more like credible intervals than box plots. Maybe a bit like the minimal box plots proposed by Edward Tufte but still I definitely would add a note to be sure the reader understands what the slabs show.

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) + 
  ggdist::stat_halfeye(
    adjust = .5,
    width = .6, 
    ## set slab interval to show IQR and 95% data range
    .width = c(.5, .95)
  ) + 
  ggdist::stat_dots(
    side = "left", 
    dotsize = .8, 
    justification = 1.05, 
    binwidth = .3
  ) +
  coord_cartesian(xlim = c(1.2, NA)) #+ coord_flip()
```

<br>

Of course, one could also add a true jitter instead of a dot plot or even a barcode. Now, I use `geom_half_dotplot()` from the `{gghalves}` package. Why? Because it comes with the possibility to add some justification which is not possible for the default layers `geom_point()` and `geom_jitter()`.

Note that the `{gghalves}` package adds also some jitter along the y axis which is far from optimal. The package provides a half–box plot alternative as well but I personally will never consider or recommend these as an option because the box plot itself can be reduced easily in width without you getting into trouble. At the same time, I believe that these “half–box plots” have an uncommon look and thus the potential to confuse readers.

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) + 
  ggdist::stat_halfeye(
    adjust = .5, 
    width = .6, 
    .width = 0, 
    justification = -.2, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .15, 
    outlier.shape = NA
  ) +
  ## add justified jitter from the {gghalves} package
  gghalves::geom_half_point(
    ## draw jitter on the left
    side = "l", 
    ## control range of jitter
    range_scale = .4, 
    ## add some transparency
    alpha = .3
  ) +
  coord_cartesian(xlim = c(1.2, NA), clip = "off") #+ coord_flip()
```

<br>

A good alternative may be to place the jittered points on top of the box plot by using geom_jitter() (or geom_point()):

```{r fig.width = 5, fig.height = 3}
ggplot(data, aes(x = group1, y = value)) + 
  ggdist::stat_halfeye(
    adjust = .5, 
    width = .6, 
    .width = 0, 
    justification = -.3, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .25, 
    outlier.shape = NA
  ) +
  geom_point(
    size = 1.3,
    alpha = .3,
    position = position_jitter(
      seed = 1, width = .1
    )
  ) + 
  coord_cartesian(xlim = c(1.2, NA), clip = "off") #+ coord_flip()
```


<br>

```{r fig.width = 5, fig.height = 5}
data |> 
  ggplot(aes(y = group1, x = value, fill = group1)) +
  # Add slab half-violin - use stat_* for raw data or geom_* for summarized data
  ggdist::stat_halfeye(
    adjust = .5,          # custom bandwidth
    height = .5,          # adjust height (use width for vertical orientation)
    .width = 0,           # remove whiskers of slab 
    point_colour = NA,    # remove points of slab
    justification = -.15, # move geom 
  ) + 
  # Add boxplot
  geom_boxplot(width = .12, outlier.shape = NA, show.legend = F) +
  # Add dots shower
  ggdist::stat_dots(
    side = "bottom",     # orientation
    binwidth = .25,      # adjust grouping (binning) of observations 
    show.legend = F
  ) + 
  ggsci::scale_fill_lancet() +
  coord_cartesian(ylim = c(1.4,2.8)) 
```




```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```



Raincloud plot that highlights the distribution of Vehicle Fuel Economy (MPG) by Engine Size (Number of Cylinders).


```{r fig.width = 5, fig.height = 5}
mpg |> 
  filter(cyl %in% c(4,6,8)) |> 
  ggplot( aes(x=factor(cyl), y=hwy, fill=factor(cyl)) ) + 
  
  # add half-violin from {ggdist}
  ggdist::stat_halfeye( 
    adjust = 0.5,         # custom bandwidth
    justification = -.2,  # move geom to the right
    .width = 0,           # remove slab interval
    point_colour = NA,   # remove points
  ) +
  # add the boxplot 
  geom_boxplot(
    width = .12,
    outlier.color = NA,   # remove outliers
    alpha = 0.5, show.legend = F
  ) + 
  # add the dot plots from {ggdist}
  ggdist::stat_dots(
    side = "bottom",      # orientation to the bottom
    justification = 1.1,  # move geom down
    binwidth = .25,
    show.legend = F
  ) +
  # make it look professional
  ggsci::scale_fill_lancet() +
  theme(
    axis.title = element_text(),
    panel.grid.major.x = element_line(),
    panel.border = element_rect(fill = NA),
    legend.title = element_text(),
    legend.position = "top"
  ) +
  labs(
    title = "Raincloud Plot",
    subtitle = "Showing the Bi-Modal Distribution of 6 Cylinder Vehicles",
    x = "Engine Size (No. of Cylinders)",
    y = "Highway Fuel Economy (MPG)",
    fill = "Cylinders"
  ) +
  coord_flip()
```


```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```


<br>

# Heatmap

```{r fig.width = 5, fig.asp=2}
diamonds |>
  mutate(cut_clarity = paste(cut, clarity)) |>
  group_by(cut_clarity,cut, clarity, color) |>
  summarise(n = n(),
            mean_price = mean(price),
            sd = sd(price),
            .groups = "drop") |> 
  
  ggplot(aes(x = color, y = cut_clarity)) +
  geom_tile(aes(fill = mean_price))
```


<br>



# Hexagonal 2d

<br>

```{r}
d <- ggplot(diamonds, aes(carat, price))
d + geom_hex() + viridis::scale_fill_viridis(option = "A", direction = -1)


# You can control the size of the bins by specifying the number of
# bins in each direction:
d + geom_hex(bins = 10) + viridis::scale_fill_viridis(option = "A", direction = -1)

d + geom_hex(bins = 30) + viridis::scale_fill_viridis(option = "B", direction = -1)


# Or by specifying the width of the bins
d + geom_hex(binwidth = c(1, 1000)) + viridis::scale_fill_viridis(option = "C", direction = -1)

d + geom_hex(binwidth = c(.1, 500)) + viridis::scale_fill_viridis(option = "D", direction = -1)
```


```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```


<br>

# ggside

<br>

Marginal plots are a way to extend your numeric data with side plots that highlight the density (density, histogram or boxplots). Marginal Distribution Plots were made popular with the seaborn jointplot() side-panels in Python. 

The trick is using the package `ggside` and `after_stat(density)`, which makes an awesome looking marginal density side panel plot.

[ggside Walkthrough](https://cran.r-project.org/web/packages/ggside/vignettes/ggside_basic_usage.html)


```{r fig.width = 7, fig.asp = .6}
library(ggside)

iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  geom_xsidedensity(aes(y=stat(density))) +
  geom_ysidedensity(aes(x=stat(density))) +
  
  theme_bw() +
  labs(title = "FacetNull")
```


<br>

### Plot types

```{r fig.width = 7, fig.asp = .6}
p <-ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species, fill = Species)) +
  geom_point() +
  theme_bw() 

#sidecol - uses Global mapping
p2 <- p +
  geom_xsidecol() +
  geom_ysidecol() +
  labs(title = "Column")

# #sidebar - uses StatCount
# p1 <- p +
#   geom_xsidebar() +
#   geom_ysidebar() +
#   labs(title = "Bar")

p2
```


```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```


<br>

```{r fig.width = 10, fig.height = 4}
df <- expand.grid(UpperCase = LETTERS, LowerCase = letters)
df$Combo_Index <- as.integer(df$UpperCase)*as.integer(df$LowerCase)
p <- ggplot(df, aes(UpperCase, LowerCase)) +
  geom_tile(aes(fill = Combo_Index)) +
  theme_bw()

# Note - Mixing discrete and continuous axis scales using xsideboxplots when the y aesthetic was
# previously mapped with a continuous variable will prevent any labels from being plotted. 
# This is a feature that will hopefully be added to ggside in the future.
p1 <- p + geom_xsideboxplot(aes(y = Combo_Index)) +
  geom_ysideboxplot(aes(x = Combo_Index)) +
  labs(title = "Boxplots")

# sideboxplots with swapped orientation
# Note - Discrete before Continuous
# If you are to mix Discrete and Continuous variables on one axis, ggplot2 prefers the discrete
# variable to be mapped BEFORE the continuous.
p2 <- ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_xsideboxplot(aes(y = Species), orientation = "y") +
  geom_point() +
  theme_bw() +
  labs(title = "Boxplots with swapped orientation")

# #Alternatively, you can recast discrete as a factor and then a numeric
# p3 <- ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species))+
#   geom_point() +
#   geom_xsideboxplot(aes(y = as.numeric(Species)), orientation = "y") +
#   geom_ysideboxplot(aes(x = as.numeric(Species)), orientation = "x") +
#   theme_bw() +
#   labs(title = "Boxplots with swapped orientation")

p1+p2
```



```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```

<br>


```{r fig.width = 10, fig.height = 4}
p <- ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species, fill = Species)) +
  geom_point()

p1 <- p +
  geom_xsidehistogram(binwidth = 0.1) +
  geom_ysidehistogram(binwidth = 0.1) +
  theme_bw() +
  theme(ggside.panel.scale=0.3) + # set side panel relative size
  labs(title = "Histogram")

p2 <- p +
  # geom_*sidehistogram uses the same aesthetics as geom_*sidebar()
  geom_xsidehistogram(aes(y = after_stat(density)), binwidth = 0.1) +
  geom_ysidehistogram(aes(x = after_stat(density)), binwidth = 0.1) +
  theme(ggside.panel.scale=0.3) + # BEFORE THEME UPDATES !
  theme_bw() +
  labs(title = "Histogram")

p1+p2
```



```{r, include=FALSE}
rm(list=ls()) # Limpa o global environment
```


<br>

```{r fig.width = 5, fig.height = 3}
ggplot(economics, aes(date, pop)) +
  geom_col() +
  geom_xsideline(aes(y = unemploy)) +
  theme_bw() +
  theme(ggside.panel.scale=0.5) + # set side panel relative size
  labs(title = "Line")
```

<br>

```{r fig.width = 8, fig.asp = 1}
ggplot(diamonds, aes(depth, table, alpha = .2)) +
  geom_point() +
  geom_ysidepoint(aes(x = price)) +
  geom_xsidepoint(aes(y = price)) +
  theme_bw() +
  theme(ggside.panel.scale = .3) +
  labs(title = "Point")
```



<br>

### With Facets

```{r fig.width = 6, fig.height = 6}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(x=Sepal.Width, y=Sepal.Length, color = Species)) +
  geom_point() + 
  theme(axis.title = element_text()) +
  
  # geom_xsidedensity(aes(y=stat(density))) +
  # geom_ysidedensity(aes(x=stat(density))) +
 
  theme_bw() +
  facet_wrap(Species~Species2, ncol = 2) +
  guides(guide_axis(check.overlap = T)) +
  labs(title = "FacetWrap")
```


```{r fig.width = 6, fig.height = 6}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # geom_xsidedensity(aes(y=stat(density))) +
  # geom_ysidedensity(aes(x=stat(density))) +
 
  theme_bw() +
  facet_grid(Species~Species2, space = "free", scale = "free_y") +
  labs(title = "FacetGrid")
```

<br>

Control how the sideFacets are handled with the ggside function

```{r}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # ggside functions
  geom_xsidedensity(aes(y=stat(density))) +
  geom_ysidedensity(aes(x=stat(density))) +
  ggside(x.pos = "bottom", y.pos = "left") +
  
  # beautify
  theme_bw() +
  labs(title = "FacetNull", 
       subtitle = "Xside placed bottom, Yside placed left")
```

<br>

When using multiple panels, you can collapse side panels to one side


```{r}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # ggside functions
  # geom_xsidedensity(aes(y=stat(density))) +
  # geom_ysidedensity(aes(x=stat(density))) +
  # ggside(collapse = "x") +
  
  # beautify
  theme_bw() +
  facet_wrap(Species~Species2, ncol = 3) +
  labs(title = "FacetWrap", subtitle = "Collapsing X side Panels")
```

<br>

Note that when collapsing panels on `FacetGrid`, the panels appear under the strips whereas on `FacetWrap` they appear above the strips. This is because `FacetWrap`, collapsing panels in the same column or row may not share the same facet variable, which would be confusing since the strip would not represent the data entirely. This is not the case with `FacetGrid` since each row or column is dictated by the facet variable.


```{r fig.width = 6, fig.height = 6}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # # ggside functions
  # geom_xsidedensity(aes(y=stat(density))) +
  # geom_ysidedensity(aes(x=stat(density))) +
  # ggside(collapse = "all") +
  # 
  # beautify
  theme_bw() +
  facet_grid(Species~Species2, space = "free", scales = "free") +
  labs(title = "FacetGrid", subtitle = "Note diferentiated groups in X only")
```

<br>

You can control which variable is mapped to the side panel:

```{r fig.width = 6, fig.height = 6}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # # ggside functions
  # geom_xsidedensity(aes(y=stat(density))) +
  # geom_ysidedensity(aes(x=stat(density), ycolor=Species2)) + # Differentiate y in x-side panel
  # ggside(collapse = "all") +
 
  theme_bw() +
  facet_grid(Species~Species2, space = "free", scales = "free") +
  labs(title = "FacetGrid", subtitle = "Differentiated groups in X and Y")
```

<br>

You can fully control the elements of the side panels separately from the main plots

```{r fig.width = 6, fig.height = 6}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # # ggside functions
  # geom_xsidedensity(aes(y=stat(density), xfill = Species), 
  #                   position = "stack")+
  # geom_ysidedensity(aes(x=stat(density), yfill = Species2), 
  #                   position = "stack") +
  # ggside(collapse = "all") +
  # scale_xfill_manual(values = c("darkred","darkgreen","darkblue")) +
  # scale_yfill_manual(values = c("black","gold")) +
  # scale_ysidex_continuous(guide = guide_axis(check.overlap = T)) + # fix overlapping axis labels
  # scale_xsidey_continuous(guide = guide_axis(check.overlap = T)) + # fix overlapping axis labels

  # beautify
  theme_bw() + 
  facet_grid(Species~Species2, space = "free", scales = "free") +
  labs(title = "FacetGrid", subtitle = "Collapsing All Side Panels")
```

<br>

You may also change the size of the side panels with ggside theme elements. 


```{r fig.width = 6, fig.height = 6}
iris |> 
  mutate(Species2 = rep(c("A","B"), 75),
         Species = as.character(Species)) |> 
  
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() + 
  
  # # ggside functions
  # geom_xsidedensity(aes(y=stat(density))) +
  # geom_ysidedensity(aes(x=stat(density))) +
  # ggside(collapse = "x", x.pos = "bottom", scales = "free_x") +
  # theme(ggside.panel.scale.x = .4,      # as percent of main plot(s)
  #       ggside.panel.scale.y = .25) +   # as percent of main plot(s)
  # scale_ysidex_continuous(guide = guide_axis(check.overlap = T)) + # fix overlapping axis labels
  # scale_xsidey_continuous(guide = guide_axis(check.overlap = T)) + # fix overlapping axis labels
  
  # beautify
  theme_bw() +
  facet_grid(Species~Species2, space = "free", scales = "free") +
  labs(title = "FacetGrid", 
       subtitle = "Collapsed X Side Panels and Adjusted Side Panel Relative Size")
```

<br>

You can choose which type of plot to put on the side panels.
You can specify ggside Axes.

You can have control over how a side axis will render. For example, when making an xside geometry, the x-axis is shared with the main panel so you can specify how the x-axis is rendered via the `scale_x_*` functions. You can use `scale_xsidey_(continuous|discrete)` functions to further specify this scale.

Similarly, you can do this for the x-axis of a yside panel with `scale_ysidex_(continuous|discrete)` functions. For all intents and purposes, these are identical to the `scale_(x|y)_*` functions but they only affect their xside or yside panel's non-shared axis.

Additionally, this allows you to mix continuous and discrete scales on the same y or x axis. For example the main panel y axis may be continuous and the side panel y axis may be discrete. Take the following example which will generate an error:

    mpg |> filter(class %in% c("compact","suv","2seater")) |> 
      ggplot(aes(displ, hwy, colour = class)) + 
      geom_point(size = 2) + 
      
      geom_xsideboxplot(aes(y =class), orientation = "y") +
      geom_ysidedensity(aes(x = after_stat(density)), position = "stack") +
      theme(ggside.panel.scale = .3)

<br>

You can provide the plot with the proper scale the panel will expect. You can use the guide argument of these new scale functions to further customize how the text is rendered, the breaks argument to control the location or visibility of the tick marks.

```{r}
mpg |> filter(class %in% c("compact","suv","2seater")) |> 
  ggplot(aes(displ, hwy, colour = class)) + 
  geom_point(size = 2) + 
  # ggside functions
  geom_xsideboxplot(aes(y =class), orientation = "y") +                  # choice of plot
  geom_ysidedensity(aes(x = after_stat(density)), position = "stack") +  # choice of plot
  theme(ggside.panel.scale = .3) + # relative size of side panel
  scale_xsidey_discrete() +        # change x side panel scale to discrete
  theme_bw() +
  scale_ysidex_continuous(         
    guide = guide_axis(angle = 90, check.overlap = T), # format the axis
    minor_breaks = NULL             # format the axis breaks
  )
```


<br>

Beautiful example:

```{r}
mpg |> filter(class %in% c("compact","suv","2seater")) |> 
  ggplot( aes(hwy, cty, color=class) ) +
  geom_point(size=2, alpha=0.3) +
  geom_smooth(aes(color=NULL), se=TRUE) +
  # ggside
  ggside::geom_xsidedensity(                    # add side density panel on top (to X)
    aes(y = after_stat(density), fill = class), # specify aesthetics
    alpha=0.5, size=1, position="identity"      # format the plot
  ) +
  ggside::geom_ysidedensity(                    # add side density panel on right (to Y)
    aes(x = after_stat(density), fill = class), # specify aesthetics
    alpha=0.5, size=1, position="identity"      # format the plot
  ) +
  theme(
    ggside.panel.scale.x = 0.4, # CONTROL THE RELATIVE SIZE OF SIDE PANELS
    ggside.panel.scale.y = 0.4
  ) +
  # beautify
  ggsci::scale_color_lancet() +
  ggsci::scale_fill_lancet() +
  theme_bw() +
  labs(
    title = "Fuel Economy by Vehicle Type",
    subtitle = "ggside density",
    x = "Highway MPG", y = "City MPG"
  ) 
```

